Fixing Vulnerabilities Hinders Maintainability

### 2019-02-02 paper/ICPC19/icpc19.tex
fixes

\author{
    Anonymou(s) Author(s)
%     \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
% \and
% \IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
}

@@

As an example, consider the refactoring of the Datagram Transport Layer Security
(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a tookit that
contains open-source implementations of the SSL and TLS cryptographic
protocols.} to address the fact that hackers could cause a Denial-of-Service
(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506
details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The following snippet presents the changes performend on the
\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
(Accessed on \today{})} by the OpenSLL developers to fix the vulnerability.

@@

- unsigned long frag_len = msg_hdr->frag_len, max_len; /*!\annotation{lst:func3}!*/
- if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
-    goto err;

@@

+ if(frag_len > dtls1_max_handshake_message_len(s))
+    goto err; /*!\annotation{lst:func4}!*/
+
  frag = dtls1_hm_fragment_new(frag_len, 0);

@@

\ref{lst:func3} Replaced by attribution in line $23$.

@@

Although limiting the size of a handshake message seems an elementary problem to
repair, the example shows that there was a significant amount of changes
performed in the codebase which produced a negative impact in the project
maintainability. Even though the developer tries to simplify the codebase with
the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it
also disrupts $2$ of the guidelines proposed by the Software Improvement Group
(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
to a potential violation of the
\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
impact of this refactoring on the OpenSSL maintainability was negative.

@@

Often security flaws require refactoring code to make software more secure.
However, there is no evidence yet of how security refactorings impact the
maintainability of open-source software. Our suspicion is that developers tend
to introduce technical debt in their software when refactoring the codebose to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability value of $607$ commits using
\emph{Better Code Hub}. The same approach is applied to a randomly generated
dataset of regular commits - baseline - that we use to understand how
maintainability evolves when security refactorings are performed versus when
they are not.

@@

There are security flaws that are more difficult to refactor than others. For
instance, introducing a mechanism to secure authentication is not so easy to
implement as fixing a cross-site scripting vulnerability with a \texttt{escape}
function. Knowing which patterns are more likely to increase maintainability
issues is one step forward to bring awareness to security engineers of what
patterns need more attention. The taxonomy of security patterns used to answer
this question is the same as the one provided by the authors of the dataset.
Maintainability is measured separately for each pattern.
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed in the following sections
and illustrated in Figure \ref{fig:met}. Aiming to answer the research questions
outlined in the previous section, in this paper, we use a dataset containing
$716$ security refactorings collected from open-source software available on
GitHub~\cite{Reis:2017:IJSSE}. Moreover, a baseline of regular commits was
randomly collected from the list of projects of the main dataset to evaluate
what is the impact of regular commits on the maintainability of open-source
software. The Software Improvement Group's\footnote{SIG's website is available
at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
https://bettercodehub.com/ (Accessed on \today{})} was used to quantify
maintainability for both security and regular commits.
%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu (2017) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools \cite{just2014mutants}. The study yielded a
dataset with $716$ test case for $16$ security patterns. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring and the snippets of code that differ from one version to
another (typically, called \textit{diff}) --- where one can easily identify the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.

@@

The test cases in the dataset are classified with one of the $16$ following
patterns, as based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top
10 of 2017~\cite{oswap:2017}: \textit{Injection}, \textit{Broken Authentication
and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access
Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure},
\textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery},
\textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs},
\textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service},
\textit{Path Traversal}, \textit{Miscellaneous}, \textit{Resource Leak} and
\textit{SHA-1 Hash Function}.

@@

\begin{table*}[h]
	\centering
	\caption{Descriptive statistics of the dataset projects}
\begin{tabular}{@{}lllllllllll@{}}
\toprule
      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
mean  & 1763.52 & 5448.74 & 401.37   & 153.33       & 14834.17 & 45.17    & 129.45   & 122973.24 & 3768.97   & 1941.61 \\
std   & 2434.03 & 6215.09 & 486.60   & 123.68       & 22234.46 & 150.15   & 189.93   & 209732.51 & 5933.16   & 3603.31 \\
min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
25\%  & 391.50  & 1581    & 117.25   & 49           & 1440.50  & 4        & 19       & 8466.75   & 313.75    & 143.25  \\
median  & 838.50  & 2836.50 & 248      & 99           & 5504.50  & 9        & 59       & 37372.50  & 1792.50   & 650     \\
75\%  & 2155    & 6828.50 & 459.50   & 261          & 18579.25 & 20       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
\end{tabular}
\label{tab:dataset}
\end{table*}

The dataset contemplates security flaws for more than 13 different languages
being PHP (39\%) and C (24\%) the most prevalent ones; see Figure \ref{fig:lang}.
To classify the applications in the dataset, we use a taxonomy for open-source
software published in an older study~\cite{7816479}. Figure \ref{fig:domain}
presents the projects domain distribution: \textit{Aplication Software} ($25$),
software that provides end-users with functional systems; \textit{Web libraries
and frameworks} ($20$); \textit{System Software} ($18$), software that provides
services and infrustructures (e.g., operating systems, servers and databases);
\textit{Software Tool} ($13$), software that supports development (e.g.,
programming languages, compilers, package managers, IDEs); and, \textit{Non-web
libraries and frameworks} ($18$), for desktop and mobile software.
%
Table \ref{tab:dataset} presents the descriptive statistics of the $94$
open-source projects involved in this study, icluding number of forks, number of
stars, number of watchers, number of contributors, number of commits, number of
branches, number of releases,  size (cf. GitHub it is the size of the whole
repository, including all of its history, in kilobytes), number of issues and pull
requests.

%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of code refactorings on
open-source software maintainability~\cite{HEGEDUS2018313} before. But, to the
best of our knowledge, no previous work has used the BCH model and focused on
evaluating the impact of security refactorings and compare with regular
refactorings as in this study. We analyze the maintainability of regular commits
(i.e., commits not related with security fixes) and use them as a baseline.

@@

Figure \ref{fig:bchrep} shows an example of the  guidelines report
provided by BCH for a project after finishing its evaluation. These are the
results after performing the OpenSSL CVE-2014-3506 vulnerability refactoring -
motivation example discussed, previously, in section \ref{sec:motivation}. This
version of OpenSSL only complies with 1 out of 10 guidelines, the \emph{Write
Clean Code} guideline.

>>>>

\author{
    Anonymou(s) Author(s)

%     \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
% \and
% \IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
}

@@

As an example, consider the refactoring of the Datagram Transport Layer Security
(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a tookit that
contains open-source implementations of the SSL and TLS cryptographic
protocols.} to address a vulnerability where an hacker could cause a Denial-of-Service
(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506
details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The following snippet presents the changes performend on the
\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
(Accessed on \today{})} by the OpenSLL developers to fix the vulnerability.

@@

- unsigned long frag_len = msg_hdr->frag_len, max_len; 
- if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
-    goto err;

@@

+ if(frag_len > dtls1_max_handshake_message_len(s))
+    goto err; /*!\annotation{lst:func4}!*/

  frag = dtls1_hm_fragment_new(frag_len, 0);

@@

@@

Although limiting the size of a handshake message seems an elementary problem to
repair, the example shows that there was a significant amount of changes
performed in the codebase which produced a negative impact in the project
maintainability. Even though the developer tries to simplify the codebase with
the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it
also disrupts $2$ of the guidelines proposed by the Software Improvement Group
(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
to a potential violation of the
\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
impact of this refactoring on the OpenSSL maintainability was negative.

@@

Often security flaws require refactoring code to make software more secure.
However, there is no evidence yet of how security refactorings impact the
maintainability of open-source software. Our suspicion is that developers tend
to introduce technical debt in their software when refactoring the codebose to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $607$ commits using
\emph{Better Code Hub}. The same approach is applied to a randomly generated
dataset of regular commits --- baseline --- that we use to understand how
maintainability evolves when security refactorings are performed versus when
they are not.

@@

There are security flaws that are more difficult to refactor than others. For
instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with an \texttt{escape}
function. Knowing which patterns are more likely to increase maintainability
issues is one step forward to bring awareness to security engineers of what
patterns need more attention. The taxonomy of security patterns used to answer
this question is the same as the one provided by the authors of the dataset~\cite{Reis:2017:IJSSE}.
Maintainability is measured separately for each pattern.
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed this section. The methodology 
comprehends the following steps, as illustrated in Figure \ref{fig:met}:
\begin{enumerate}
	\item Extract relevant data from a dataset containing
$716$ security refactorings collected from open-source software available on
GitHub~\cite{Reis:2017:IJSSE}.
	\item A baseline of regular commits was
randomly collected from the list of projects of the main dataset to evaluate the impact of regular commits on the maintainability of open-source
software.
\item Use the Software Improvement Group's\footnote{SIG's website is available
at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
https://bettercodehub.com/ (Accessed on \today{})} to quantify
maintainability for both security and regular commits.
\end{enumerate}

%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu (2017) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools \cite{just2014mutants}. The study yielded a
dataset with $716$ test cases for $16$ security patterns. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring, and the snippets of code that differ from one version to
another (typically, called \textit{diffs}) --- where one can easily review the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.

@@

The test cases in the dataset are classified with one of the $16$ following
patterns: \textit{Injection}, \textit{Broken Authentication
and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access
Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure},
\textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery},
\textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs},
\textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service},
\textit{Path Traversal}, \textit{Miscellaneous}, \textit{Resource Leak} and
\textit{SHA-1 Hash Function}. Some of the patterns are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 2017~\cite{oswap:2017}

@@

\begin{table*}[h]
	\centering
	\caption{Descriptive statistics of the dataset projects}
\begin{tabular}{@{}rrrrrrrrrrr@{}}
\toprule
      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
mean  & 1763.52 & 5448.74 & 401.37   & 153.33       & 14834.17 & 45.17    & 129.45   & 122973.24 & 3768.97   & 1941.61 \\
std   & 2434.03 & 6215.09 & 486.60   & 123.68       & 22234.46 & 150.15   & 189.93   & 209732.51 & 5933.16   & 3603.31 \\
min   & 1.00       & 3.00       & 1.00        & 0.00            & 103.00      & 1.00        & 0.00        & 108.00       & 0.00         & 0.00       \\
25\%  & 391.50  & 1581.00    & 117.25   & 49.00           & 1440.50  & 4.00        & 19.00       & 8466.75   & 313.75    & 143.25  \\
median  & 838.50  & 2836.50 & 248.00      & 99.00           & 5504.50  & 9.00        & 59.00       & 37372.50  & 1792.50   & 650.00     \\
75\%  & 2155.00    & 6828.50 & 459.50   & 261.00          & 18579.25 & 20.00       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
max   & 16366.00   & 31841.00   & 3446.00     & 413.00          & 114378.00   & 1227.00     & 1114.00     & 995790.00    & 33970.00     & 19329.00   \\
Total & 165771.00  & 512182.00  & 37729.00    & 14413.00        & 1394412.00  & 4246.00     & 12168.00    & 11559485.00  & 354283.00    & 182511.00  \\ \bottomrule
\end{tabular}
\label{tab:dataset}
\end{table*}

The dataset contemplates security flaws for more than 13 different languages
being PHP (39\%) and C (24\%) the most prevalent ones; see Figure \ref{fig:lang}.
To classify the applications in the dataset, we use a taxonomy for open-source
software from previous work~\cite{7816479}. Figure \ref{fig:domain}
presents the projects domain distribution: \textit{Aplication Software} ($25$),
software that provides end-users with functional systems; \textit{Web libraries
and frameworks} ($20$); \textit{System Software} ($18$), software that provides
services and infrustructures (e.g., operating systems, servers and databases);
\textit{Software Tool} ($13$), software that supports development (e.g.,
programming languages, compilers, package managers, IDEs); and, \textit{Non-web
libraries and frameworks} ($18$), for desktop and mobile software.
%
Table \ref{tab:dataset} presents the descriptive statistics of the $94$
open-source projects involved in this study, including the number of forks, the number of
stars, the number of watchers, the number of contributors, the number of commits, the number of
branches, the number of releases,  size (cf. GitHub it is the size of the whole
repository, including all of its history, in kilobytes), the number of issues, and the
number of pull requests.

%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of regular refactorings on
open-source software maintainability~\cite{HEGEDUS2018313}. However, 
there is no previous work focused on comparing the impact of security 
refactorings with regular refactorings on maintainability. 
We analyze the maintainability of regular commits
(i.e., commits not related with security fixes) and use them as a baseline.

@@

Figure \ref{fig:bchrep} shows an example of the report
provided by BCH for a project after finishing its evaluation. The examples
refers to the OpenSSL CVE-2014-3506 vulnerability refactoring ---
as described as motivating example in Section \ref{sec:motivation}. This
version of OpenSSL only complies with 1 out of 10 guidelines: \emph{Write
Clean Code}.
